---
title: "SMPA - Grouping Assignment"
author: "Riya Sharma"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)

#load data
payments <- readRDS("data/trump_campaign_payments.rds")

```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  Which organization had the highest disbursement?

2.  Which type of disbursement (disbursement_description) is most popular?

3.  Are most entities individuals or organizations?

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

-   To answer this question, I first called the payments dataset and used the tidyverse pipe operator to connect the group_by() command. I then grouped by the report year (2017 and 2018) and used summarize to create a category called amount_spent, which added together the disbursements from each year with the sum() function and displayed them in a resulting tibble. The results show that $16,888,149	was spent in 2017 and $37,406,905	was spent in 2018.

```{r}

payments %>%
  group_by(report_year) %>%
  summarize(amount_spent = sum(disbursement_amount))





```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

- In order to tally by month and distinguish by year, I grouped the payments dataset by the month and year variables and summarized by the total disbursement amount using the sum() function. Lastly, I arranged the amount_spent in descending order to discover that spending was highest in October 2018 at $9,738,567.60. 

```{r}

payments %>%
  group_by(disbursement_month, disbursement_year) %>%
  summarize(amount_spent = sum(disbursement_amount)) %>%
  arrange(desc(amount_spent))

```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?

- In order to check for the 2020 election cycle specifically, I filtered the election_cycle column for the year 2020. Then, I grouped the dataset by recipient and summarized by the total disbursement amount using the sum() function for the disbursement_amount variable and naming it amount_paid. Lastly, I arranged the amount_paid in descending order to discover that the recipient paid the most overall was American Made Media Consultants, LLC at $9,772,583.64.

```{r}

payments %>%
  filter(election_cycle == 2020) %>%
  group_by(recipient_name) %>%
  summarize(amount_paid = sum(disbursement_amount)) %>%
  arrange(desc(amount_paid))

```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

- I found that they were paid the most for "Placed Media". I first called the payments dataset and used the tidyverse pipe operator to connect the group_by() command. Then, I once again filtered by the 2020 election cycle and grouped by recipient_name. I used summarize() to sort by the disbursement_description and disbursement_amount and arranged the disbursement_amount variable in descending order so the highest payment would show up first. 

```{r}

payments %>%
  filter(election_cycle == 2020) %>%
  group_by(recipient_name) %>%
  summarize(disbursement_description, disbursement_amount) %>%
  arrange(desc(disbursement_amount))




```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

- I first called the payments dataset and used the tidyverse pipe operator to connect the group_by() command. Then, I grouped by disbursement_description and used summarize() to sort by the total number of times the campaign spent money on a specific type of disbursement. Finally, I arranged the total variable in descending order so the most common payment would show up first. The most common description was "Merchant Fees" with a total of 1002 disbursements. 

```{r}

payments %>%
  group_by(disbursement_description) %>%
  summarize(total = n()) %>%
  arrange(desc(total))




```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

- I first called the payments dataset and used the tidyverse pipe operator to connect the filter() function so I could filter by the "LEGAL CONSULTING" description. Then, I grouped by recipient_name and used summarize() to sort by the total number of times the campaign spent money on a specific type of disbursement and paid, which sums the disbursements for each recipient. Finally, I arranged the paid variable in descending order so the biggest, highest paid legal consultants would show up first. The most highly paid consultants were JONES DAY; LAROCCA HORNIK ROSEN GREENBERG & BLAHA LLP; KASOWITZ, BENSON, TORRESS & FRIEDMAN; THE LAW OFFICES OF ALAN S. FUTERFAS; and MCDERMOTT WILL & EMERY.  

```{r}

payments %>%
  filter(disbursement_description == "LEGAL CONSULTING") %>%
  group_by(recipient_name) %>%
  summarize(total = n(), paid = sum(disbursement_amount)) %>%
  arrange(desc(paid))




```

Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

- I first called the payments dataset and used the tidyverse pipe operator to connect the filter() function so I could filter by the "INDIVIDUAL" entity type. Then, I grouped by recipient_name and used summarize() to sort by amount_paid, which sums the disbursements for a specific individual. Finally, I arranged the amount_paid variable in descending order so the biggest, highest paid individuals would show up first. The most highly paid individuals were GLASSNER, MICHAEL; DOLLMAN, SEAN; CENTINELLO, DARREN; CARR, CHRISTOPHER; and PENCE, JOHN.  

```{r}

payments %>%
  filter(entity_type_desc=="INDIVIDUAL") %>%
  group_by(recipient_name) %>%
  summarize(amount_paid = sum(disbursement_amount)) %>%
  arrange(desc(amount_paid))


```

For the individual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

- I first called the payments dataset and used the tidyverse pipe operator to connect the filter() function so I could filter by the highest paid recipient ("GLASSNER, MICHAEL") from the question above. Then, I grouped by recipient_name and used summarize() to sort by disbursement_amount and the disbursement_description. Finally, I arranged the disbursement_amount variable in descending order so the top, most expensive categories would show up first. The top payment category was payroll for Michael Glassner. 

```{r}

payments %>%
  filter(recipient_name=="GLASSNER, MICHAEL") %>%
  group_by(recipient_name) %>%
  summarize(disbursement_amount, disbursement_description) %>%
  arrange(desc(disbursement_amount))





```
